Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-22 model.tex
px to p in model.tex

A random-walk predictor simply uses the last observed measurement as
the forecast: that is, the value $px_i$ at time $i$ is calculated
using the relation $$px_i = x_{i-1}$$ The prediction strategy that we
refer to using the term ``\naive'' averages the prior observations to
generate the forecast: $$px_i = \sum_{j=1}^{i-1}\frac{x_j}{i-1}$$
While both of these methods are simplistic, they are not without
merit.  For a time series near the high end of the complexity
spectrum---i.e., one that possesses very little predictive
structure---these two methods can actually be the best choice.  In
forecasting currency exchange rates, for instance, sophisticated
econometrics-based prediction models fail to consistently outperform
the random-walk method~\cite{rwMeese,rwCCE}.  These signals are
volatile, noisy, and possess very little predictive structure, but
they do not---on the average---vary a great deal, so the random-walk
method's strategy of simply guessing the last known value is not a bad
choice.  If a signal has a unimodal distribution with low variance,
the \naive ~prediction strategy will perform quite well---even if the
signal is highly complex---simply because the mean is a good
approximation of the future behavior.  Moreover, the \naive
~prediction strategy's temporal average effects a low-pass filtering
operation, which effectively mitigates the complexity in signals with
very little predictive structure.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% here is the old version of this material, for use in other documents...
%
%Perhaps the simplest way to capture and exploit the structure of data
%is to fit a hyperplane to the dataset and then use that it to forecast
%new data points.  The roots of this date back to Yule's 1927 invention
%of the autoregressive schema~\cite{weigend93}, which forecasts the
%next time step through a weighted average of past observations: $$px_i
%= \phi(B^p)x_{i}$$ where $\phi(\cdot)$ is a polynomial of degree $p$
%that is fit to the past $p+1$ [[???]] values of $x_i$ via the
%backshift operator $B$ \footnote{I changed the superscript to $k$ to
%  make this definition general---not connected to any of the terms
%  that have specific meanings later in this section.}:
%$$B^k x_i = x_{i-k}$$
%%
%with $k=1...p$.  To account for noise in the data, one can add a
%so-called ``moving average'' term to the model:
%$\theta(B^q)\epsilon_i$, where $\theta(\cdot)$ is a polynomial of
%degree $q$ and $\epsilon$ is white noise with the same mean and
%variance as the data.
%% if you leave this in, you'll need to explain a lot about the data
%% and the calculation
%% $\{\epsilon_i\}\sim WN(0,\sigma^2)$.  
%To remove nonstationarities in the data, one can detrend it using a
%differencing operation: $(1-B^d)x_i$.  
%
%A strategy that incorporates all three of these features is called a
%\emph{nonseasonal ARIMA} model of order $(p,d,q)$:
% $$\phi(B^p)(1-B^d) x_i = \theta(B^q)\epsilon_i$$
%%
%{\color{red} You had superscripts on some of the $B$s and not on
%  others.  I added them but I may have gotten them wrong.  I also
%  changed some $(1-B)^d$s to $(1-B^d)$s to make things consistent with
%  the first use of that construction.  Please make sure things are
%  consistent, here and throughout this section.}  Here, $p$, $d$ and
%$q$ correspond to the orders of the autoregressive, detrending, and
%moving average terms, respectively.
%
%If periodic structure is present in the data, a \emph{seasonal ARIMA}
%model of order $(p,d,q)(P,D,Q)$ can be a good choice:
%%
%$$\Phi(B^P)\phi(B^p)(1-B^m)^D(1-B^d) x_i =
%\Theta(B^Q)\theta(B^q)\epsilon_i$$
%%
%Here $\Phi(\cdot)$ and $\Theta(\cdot)$ are polynomials of degree $P$
%and $Q$, respectively [[which accomplish what purpose?  why?  how?
%    where did $D$ come from and what does it do?]].  $m$ is the
%seasonal frequency [[shouldn't this be ``period''?]].  [[Explain why
%    one should choose $P=Q=m$, since that's what the equation below
%    implicitly does.]]  To use a seasonal ARIMA model, one first uses
%the detrending term to remove nonstationarity: $$\hat{x_i} =
%(1-B^m)^D(1-B^d) x_{i}$$ and then creates the forecast $px_i$ by
%evaluating $$\Phi(B^m)\phi(B^p)\hat{X_i} =
%\Theta(B^m)\theta(B^q)\epsilon_i$$
% ...down to here
%%%%%%%%%%%%%%%%%%%%%%%%

A simple and yet powerful way to capture and exploit the structure of
data is to fit a hyperplane to the dataset and then use that it to
make predictions.  The roots of this approach date back to Yule's 1927
invention of the autoregressive schema~\cite{weigend93}, which
forecasts the next time step through a weighted average of past
observations: $$px_i = \sum_{j=1}^{i-1} a_j x_j$$ The weighting
coefficients $a_i$ are computed using a polynomial of degree $p$ that
is fit to the past $p+1$ values of $x_i$.  To account for noise in the
data, one can add a so-called ``moving average'' term to the model; to
remove nonstationarities, one can detrend the data using a
differencing operation.  A strategy that incorporates all three of
these features is called a \emph{nonseasonal ARIMA model}.  If
evidence of periodic structure is present in the data, a
\emph{seasonal ARIMA model}, which adds a sampling operation that
filters out periodicities, can be a good choice.

@@

      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colLMAForecast.png}
    \caption{\col \\ LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccLMAForecast.png}
    \caption{\gcc \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}  
        \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast.png}
    \caption{\svdfive \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   \caption{Predicted ($px_j$) versus true values ($c_j$) for
     forecasts of \col, \gcc, and \svdfive.  On this type of plot, a
     perfect prediction lies exactly on the diagonal.
%
{\color{red} Joshua, what would you think of adding a row of
  random-walk plots at the top?  The letters in the subcaptions---(a)
  (b) etc.---feel clunky here.  Finally, we should eventually change
  the label of the vertical axes on these plots to be whatever symbol
  we choose for the prediction.}
%
}
\label{fig:forecast-example}
  %\end{subfigure}
\end{figure*} 
In these images, the vertical axis is the prediction $px_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  [[We should comment on what a noise ball means on such a
    plot, as well as what a horizontal line means.]]

@@

 \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.0610$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
  \hline
  \end{tabular}
  \end{center}
 \label{default}
 \caption{MASE scores and weighted permutation entropies for all eight
   processes studied in this paper}
 \label{tab:error}
  \end{table*}%
Comparing these with the geometry of the plots in
Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA $MASE$ score for the \col signals
was $0.050$, for instance, while ARIMA scored much worse ($0.599$).
That is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while ARIMA only outperformed random walk by a
factor of 1.7 [[is this right?  there was a ``2'' here, but 1/0.599 is
    closer to 1.7]].  This is in accordance with the visual appearance
of the corresponding images in Figure~\ref{fig:forecast-example}.
While the correspondence between $MASE$ score and the visual
appearance of a $px$ versus $c$ plot is clear cut in this case, that
is not always so.  The plots of LMA predictions of \col and \svdfive
both lie near the diagonal, for instance, but the corresponding $MASE$
scores were very different: $0.050$ for \col and $0.718$ for \svdfive
(resp., 20 and 1.4 times better than random-walk forecasts of the same
signals).  This is a result of the normalization in the $MASE$ score.
Recall from Section~\ref{sec:simple} that the random-walk method
performs especially poorly on signals that oscillate rapidly.  \col
fits this description, so the random-walk error on this signal---the
denominator of the three $MASE$ scores in the first row of the
Table---is pathologically high, which skews those scores down.
Random-walk prediction is very effective for the\svdfive signal, on
the other hand, so the denominator is small and the $MASE$ scores are
skewes upwards.  Normalization is a two-edged sword here.  It
facilitates comparisons across signals of different types and lengths,
but its particular pathologies must be taken into account, as
discussed at more length in Section~\ref{sec:results}.

>>>>

A random-walk predictor simply uses the last observed measurement as
the forecast: that is, the value $p_i$ at time $i$ is calculated
using the relation $$p_i = x_{i-1}$$ The prediction strategy that we
refer to using the term ``\naive'' averages the prior observations to
generate the forecast: $$p_i = \sum_{j=1}^{i-1}\frac{x_j}{i-1}$$
While both of these methods are simplistic, they are not without
merit.  For a time series near the high end of the complexity
spectrum---i.e., one that possesses very little predictive
structure---these two methods can actually be the best choice.  In
forecasting currency exchange rates, for instance, sophisticated
econometrics-based prediction models fail to consistently outperform
the random-walk method~\cite{rwMeese,rwCCE}.  These signals are
volatile, noisy, and possess very little predictive structure, but
they do not---on the average---vary a great deal, so the random-walk
method's strategy of simply guessing the last known value is not a bad
choice.  If a signal has a unimodal distribution with low variance,
the \naive ~prediction strategy will perform quite well---even if the
signal is highly complex---simply because the mean is a good
approximation of the future behavior.  Moreover, the \naive
~prediction strategy's temporal average effects a low-pass filtering
operation, which effectively mitigates the complexity in signals with
very little predictive structure.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% here is the old version of this material, for use in other documents...
%
%Perhaps the simplest way to capture and exploit the structure of data
%is to fit a hyperplane to the dataset and then use that it to forecast
%new data points.  The roots of this date back to Yule's 1927 invention
%of the autoregressive schema~\cite{weigend93}, which forecasts the
%next time step through a weighted average of past observations: $$p_i
%= \phi(B^p)x_{i}$$ where $\phi(\cdot)$ is a polynomial of degree $p$
%that is fit to the past $p+1$ [[???]] values of $x_i$ via the
%backshift operator $B$ \footnote{I changed the superscript to $k$ to
%  make this definition general---not connected to any of the terms
%  that have specific meanings later in this section.}:
%$$B^k x_i = x_{i-k}$$
%%
%with $k=1...p$.  To account for noise in the data, one can add a
%so-called ``moving average'' term to the model:
%$\theta(B^q)\epsilon_i$, where $\theta(\cdot)$ is a polynomial of
%degree $q$ and $\epsilon$ is white noise with the same mean and
%variance as the data.
%% if you leave this in, you'll need to explain a lot about the data
%% and the calculation
%% $\{\epsilon_i\}\sim WN(0,\sigma^2)$.  
%To remove nonstationarities in the data, one can detrend it using a
%differencing operation: $(1-B^d)x_i$.  
%
%A strategy that incorporates all three of these features is called a
%\emph{nonseasonal ARIMA} model of order $(p,d,q)$:
% $$\phi(B^p)(1-B^d) x_i = \theta(B^q)\epsilon_i$$
%%
%{\color{red} You had superscripts on some of the $B$s and not on
%  others.  I added them but I may have gotten them wrong.  I also
%  changed some $(1-B)^d$s to $(1-B^d)$s to make things consistent with
%  the first use of that construction.  Please make sure things are
%  consistent, here and throughout this section.}  Here, $p$, $d$ and
%$q$ correspond to the orders of the autoregressive, detrending, and
%moving average terms, respectively.
%
%If periodic structure is present in the data, a \emph{seasonal ARIMA}
%model of order $(p,d,q)(P,D,Q)$ can be a good choice:
%%
%$$\Phi(B^P)\phi(B^p)(1-B^m)^D(1-B^d) x_i =
%\Theta(B^Q)\theta(B^q)\epsilon_i$$
%%
%Here $\Phi(\cdot)$ and $\Theta(\cdot)$ are polynomials of degree $P$
%and $Q$, respectively [[which accomplish what purpose?  why?  how?
%    where did $D$ come from and what does it do?]].  $m$ is the
%seasonal frequency [[shouldn't this be ``period''?]].  [[Explain why
%    one should choose $P=Q=m$, since that's what the equation below
%    implicitly does.]]  To use a seasonal ARIMA model, one first uses
%the detrending term to remove nonstationarity: $$\hat{x_i} =
%(1-B^m)^D(1-B^d) x_{i}$$ and then creates the forecast $p_i$ by
%evaluating $$\Phi(B^m)\phi(B^p)\hat{X_i} =
%\Theta(B^m)\theta(B^q)\epsilon_i$$
% ...down to here
%%%%%%%%%%%%%%%%%%%%%%%%

A simple and yet powerful way to capture and exploit the structure of
data is to fit a hyperplane to the dataset and then use that it to
make predictions.  The roots of this approach date back to Yule's 1927
invention of the autoregressive schema~\cite{weigend93}, which
forecasts the next time step through a weighted average of past
observations: $$p_i = \sum_{j=1}^{i-1} a_j x_j$$ The weighting
coefficients $a_i$ are computed using a polynomial of degree $p$ that
is fit to the past $p+1$ values of $x_i$.  To account for noise in the
data, one can add a so-called ``moving average'' term to the model; to
remove nonstationarities, one can detrend the data using a
differencing operation.  A strategy that incorporates all three of
these features is called a \emph{nonseasonal ARIMA model}.  If
evidence of periodic structure is present in the data, a
\emph{seasonal ARIMA model}, which adds a sampling operation that
filters out periodicities, can be a good choice.

@@

      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/colLMAForecast.png}
    \caption{\col \\ LMA}
    \label{fig:colLMA}
  \end{subfigure}
      \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/gccLMAForecast.png}
    \caption{\gcc \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}  
        \begin{subfigure}{0.6\columnwidth}
    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast.png}
    \caption{\svdfive \\ LMA}
    \label{fig:gccLMA}
  \end{subfigure}
  %\begin{subfigure}{0.5\textwidth}
  %  \includegraphics[width=1.0\textwidth]{figs/LMA_vs_ARIMA}
   \caption{Predicted ($p_j$) versus true values ($c_j$) for
     forecasts of \col, \gcc, and \svdfive.  On this type of plot, a
     perfect prediction lies exactly on the diagonal.
%
{\color{red} Joshua, what would you think of adding a row of
  random-walk plots at the top?  The letters in the subcaptions---(a)
  (b) etc.---feel clunky here.  Finally, we should eventually change
  the label of the vertical axes on these plots to be whatever symbol
  we choose for the prediction.}
%
}
\label{fig:forecast-example}
  %\end{subfigure}
\end{figure*} 
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while ARIMA
does not.  [[We should comment on what a noise ball means on such a
    plot, as well as what a horizontal line means.]]

@@

 \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.0610$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
  \hline
  \end{tabular}
  \end{center}
 \label{default}
 \caption{MASE scores and weighted permutation entropies for all eight
   processes studied in this paper}
 \label{tab:error}
  \end{table*}%
Comparing these with the geometry of the plots in
Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA $MASE$ score for the \col signals
was $0.050$, for instance, while ARIMA scored much worse ($0.599$).
That is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while ARIMA only outperformed random walk by a
factor of 1.7 [[is this right?  there was a ``2'' here, but 1/0.599 is
    closer to 1.7]].  This is in accordance with the visual appearance
of the corresponding images in Figure~\ref{fig:forecast-example}.
While the correspondence between $MASE$ score and the visual
appearance of a $p$ versus $c$ plot is clear cut in this case, that
is not always so.  The plots of LMA predictions of \col and \svdfive
both lie near the diagonal, for instance, but the corresponding $MASE$
scores were very different: $0.050$ for \col and $0.718$ for \svdfive
(resp., 20 and 1.4 times better than random-walk forecasts of the same
signals).  This is a result of the normalization in the $MASE$ score.
Recall from Section~\ref{sec:simple} that the random-walk method
performs especially poorly on signals that oscillate rapidly.  \col
fits this description, so the random-walk error on this signal---the
denominator of the three $MASE$ scores in the first row of the
Table---is pathologically high, which skews those scores down.
Random-walk prediction is very effective for the\svdfive signal, on
the other hand, so the denominator is small and the $MASE$ scores are
skewes upwards.  Normalization is a two-edged sword here.  It
facilitates comparisons across signals of different types and lengths,
but its particular pathologies must be taken into account, as
discussed at more length in Section~\ref{sec:results}.
